Next Trippy Structures\thanks{Supported by organization x.}

### 2019-07-10 paper.tex
Nieves-Tirso review

\usepackage{enumerate}
\usepackage{marginnote}
\usepackage[linesnumbered,commentsnumbered,boxed]{algorithm2e}
\usepackage{amsmath}
\usepackage{threeparttable}
\usepackage{multirow}
\usepackage[margin=0pt,skip=1pt]{subcaption}

\newcommand{\acumm}{T-Matrices} % T is for Total!
\newcommand{\acummname}{\mathsf{Trip-Matrices}}
\newcommand{\ctr}{XCTR}
\newcommand{\ctrname}{\mathsf{eXtended~Compact~Trip~Representation}}
\newcommand{\ttctr}{TTCTR}
\newcommand{\ctrq}{XCTRc}
\newcommand{\ttctrq}{TTCTRc}
\newcommand{\ttctrname}{\mathsf{Topology\&Trip-aware~Compact~Trip~Representation}}

\begin{document}
%
\title{Next Trippy Structures\thanks{Funded in part by European Union`s Horizon 2020 research and innovation programme
under the Marie Sklodowska-Curie grant agreement No 690941 (project
BIRDS). The Spanish group is also partially funded by Xunta de Galicia/FEDERUE
[CSI: ED431G/01; GRC: ED431C 2017/58; GEMA:IN852A
2018/14 and Innovapeme: IN848D-2017-2350417]; by MINECO-AEI/FEDER-UE
[Datos 4.0: TIN2016-78011-C4-1-R; and ETOMERDFD3:
TIN2015-69951-R]; and by FPI Program [BES-C-2017-0085]. M. A. Rodr´ıguez is
partially funded by Fondecyt-Conicyt grant number 1170497 and by the Millennium
Institute for Foundational Research on Data. }}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Nieves R. Brisaboa\inst{1} \and
Antonio Fari\~na\inst{1}\and
Daniil Galaktionov\inst{1,2} \and Tirso V. Rodeiro\inst{1} \and M. Andrea Rodr\'iguez\inst{3,4}}
%
\authorrunning{N. Brisaboa et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Universidade da Coru\~na, Laboratorio de Bases de Datos, A Coru\~na, Spain  \email{\{brisaboa, fari, d.galaktionov, tirso.varela.rodeiro\}@udc.es}  \and 
Enxenio S.L., A Coru\~na, Spain
\and
Computer Science Department, Universidad de Concepci\'on, Concepci\'on, Chile \and Millennium Institute for Foundational Research on Data, Santiago, Chile
\email{andrea@udec.cl }\\
 }
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
The abstract should briefly summarize the contents of the paper in
15--250 words.

@@

A practical representation for this information that supports efficient indexing would have numerous possible applications, such as transportation companies interested in studying the ridership \cite{tu2018spatial} or improve the efficiency of the network, such as in \cite{weng2018mining}. Another application that is enabled by such analysis is the targeted advertising \cite{zhang2017targeted}, as being able to predict travel patterns is extremely useful for an advertiser.

@@

\subsubsection{Outline}
This work is structured as follows. We start with a detailed review of related works in Section~\ref{sec:rw}, where we discuss the state of the art for trajectory indexing, mining and the underlying structures that our representations depend on. It is followed by Section~\ref{sec:model}, where we explain our design for a data model that can support relevant queries over transportation networks. In Section~\ref{sec:ps} we detail the composition of our representations and analyze their time efficiency for the targeted queries, followed by a practical analysis in Section~\ref{sec:exp}, where we evaluate the performance on a real transportation network. Finally, the conclusions can be found in Section~\ref{sec:conc}.

\section{Related works}
\label{sec:rw}
\subsection{Trajectory indexing}
Literature on spatial trajectory indexing can be categorized by the nature of the trajectories: they can be either constrained to a network or in free space (often called moving objects). While there are well-known queries for indexes that work on moving objects in free space \cite{DBLP:conf/vldb/PfoserJT00}, the network constrained trajectory indexes cover more diverse querying needs, as different networks involve different kinds of challenges (as in vehicular road network vs public transportation network), and also because the intended application may vary (analyzing trip-chaining patterns vs number of passengers within a time window). A comprehensive review on indexing methods can be found in \cite[Chapter 4]{DBLP:books/sp/PelekisT14}, to which we shall expand in the rest of this section to mention the most notable indexes and some new developments.

\subsubsection{Free trajectory indexing}
Several adaptations of the {\em R-Tree} \cite{DBLP:conf/sigmod/Guttman84} are widely used for the indexing of moving objects. The most common approach is to integrate the temporal dimension in the R-Tree itself, as found in the {\em STR-Tree} and the {\em TB-Tree} from \cite{DBLP:conf/vldb/PfoserJT00}.
Another common approach is to compliment the R-Tree with another similar structure, as has been done for the {\em MV3R-tree}~\cite{DBLP:conf/vldb/PapadiasT01},
where an Historical {\em R-Tree}~\cite{nascimento1998towards} to partition on the temporal dimension.

@@

Recently a framework based on Apache Spark was developed called {\em UlTraMan}~ \cite{ding2018ultraman}, that supports different kinds of partitioning schemes for large collections of trajectories to answer range, kNN or aggregation queries, allowing to repartition the dataset to maximize the query time efficiency for a given query type. Although UlTraMan has been tested over network constrained trajectories, it does not appear to be exploiting network information in any way, hence its inclusion in this category.

In the field of compact data structures, an index called {\em GraCT}~\cite{brisaboa2019gract} has been developed, based snapshots at regular time intervals using {\em k$^2$-Trees}~\cite{brisaboa2009k} and movement logs for individual trajectories to which grammar compression is applied with {\em RePair}~\cite{larsson2000off}. Because of this, GraCT is a self-indexed compact representation that supports spatio-temporal range and nearest-neighbor queries, as well as allowing for direct access to the trajectory information.

\subsubsection{Network constrained trajectory indexing}
There are also numerous approaches that use R-Tree based indexes for trajectories that are constrained to an underlying network, aiming to decrease the redundancy in the representation by separating the representation into levels. Examples of such indexes include the {\em FNR-Tree}~\cite{DBLP:conf/ssd/Frentzos03}, where the network elements are indexed with a 2D R-Tree and 1D R-Trees to index time intervals of moving objects along the edges of the network. Some of the limitations of the FNR-Tree are addressed in \cite{DBLP:journals/geoinformatica/AlmeidaG05}, where the authors propose the {\em MON-Tree}, where the moving objects are indexes in two dimensional R-Trees (the dimensions being edge position vs time). More recent alternatives, such as the {\em TMN-Tree}~\cite{chang2010tmn}, integrate {\em B$^+$-Trees}, which have proven to be more space and time efficient for indexing the temporal dimension. Alternatively, in \cite{rivera2018faster} a compact representation of time intervals is proposed using two bitvectors, that can be combined with these R-Tree based indexes to increase the efficiency for temporal filtering. Refer to \cite{john2017performance} for a quick comparison of these R-Tree based indexes.

@@

Another relevant representation is described in \cite{DBLP:conf/gis/KroghPTT14}. Their proposed index, {\em NETTRA}, was designed to efficiently solve a specific kind of query called {\em Strict Path Queries (SPQ)}, built on a traditional RDBMS with $B^+$-Tree indexes. Another distinctive feature of NETTRA is its treatment of network constrained trajectories as textual information, which allows to apply string matching techniques such as fingerprinting functions to determine what trajectories have similar paths on their traversed edges. This close equivalence between trajectories and strings has been further exploited by Geodabs~\cite{chapuis2018geodabs}, where both the spatial distribution and sequence information are taken into account for finding trajectories by similarity with fingerprinting. These text-based approaches are sometimes tangled with works on the topic of semantic trajectories such as \cite{al2017semantictraj}.

@@

A recent compact data structure named {\em CiNCT} has been proposed in \cite{koide2018cinct}, where trajectories are encoded into a string, that is used to build an FM-index \cite{DBLP:conf/focs/FerraginaM00} with a Huffman-shaped Wavelet Tree \cite{ferragina2009compressed}. To further save space, the string is constructed with relative movement labels instead of absolute edges, with an auxiliary structure that represents a network graph built from the input trajectories themselves. While this structure excels at pattern-matching and path extraction in vehicular networks (such as the streets of a city), it cannot be applied to our problems in public transportation networks.

\subsection{Trajectory data mining}
Currently there are several known techniques that would allow to collect data regarding users' trips over a public transportation network. Numerous works exist where those trajectories are mined from the transactions of smart cards \cite{bhaskar2015passenger,wang2014aggregated}. This can be complemented with information derived from GPS devices, as shown in \cite{ma2014development}. Alternatively, reliable trajectories may be extracted relying on positioning obtained from cellular networks, as proven by \cite{liu2017exploring}.

@@

This summarized review proves that, although we were not able to access real data from a public transportation company for this work, such curated information about users' trips can indeed be obtained and used for our proposed representations.

\subsection{Underlying structures and algorithms}
\subsubsection{Summed Area Tables}
The Summed Area Tables were first introduced in computer graphics \cite{crow1984summed} to speed up the mipmapping process. In this work we use this approach in order to manage aggregated data about user trajectories based on the idea that an amount of commuters share the same ride (with all its implications: times, trajectories...).

\subsubsection{Bitvectors}
\label{sec:bit}
A great amount of works in Compressed Data Structures revolve around bitvectors, both compressed and uncompressed. For a bitvector $B[1,n]$ the following operations are expected to be supported:

\begin{itemize}
    \item $rank_1(B,i)$ returns the number of set bits in $B[1..i]$. Alternatively, $rank_0(B,i) = i - rank_1(B,i)$ and also $B[i] = rank_1(B,i-1) - rank_1(B,i)$.
    \item $select_1(B,i)$ returns the position in $1..n$ where the $i$th 1 occurs. Therefore, $rank_1(B,select_1(B,i)) = i$.
\end{itemize}

@@

\subsubsection{Compressed Suffix Array (CSA)}
\label{sec:csa}
One common structure to count or locate all the occurrences of a pattern $P[1,m]$ in a string $S[1,n]$ of symbols from an alphabet $\Sigma$ is the Suffix Array (SA) \cite{MM93}. The SA lexicographically sorts all the suffixes $\{S[i..n] : 1 \leq i \leq n\}$ and builds an array $A[1,n]$ with their indices, holding that $S[A[i]..n] \prec S[A[i+1]..n]$. This allows to find out all the occurrences of a substring $P[1,m]$ in $O(m \log n)$ time with binary searches.

A simple enhancement to avoid storing the original string $S$ is to align $A$ with a bitvector $D[1,n]$ where $D[i]=1~\Leftrightarrow~S[A[i-1]] \prec S[A[i]]$. That is, $D$ marks with a 1 all the indexes $i$ of $A$ where the starting symbol of that suffix $S[A[i]]$ has a larger lexicographical value than the one from the previous suffix $S[A[i-1]]$. With $D$, keeping $S$ is not longer needed as $S[A[i]] = rank_1(D,i)$.
%, where $rank_1(D,i)$ returns the number of set bits in $D[1..i]$.

@@

\section{Our model for compact trip representation}
\label{sec:model}
In our previous structures introduced in \cite{brisaboa2018new}, we propose a work-in-progress model to represent the demand information for a public transportation network, as seen in the Figure~\ref{fig:er}.

@@

This approach allows us to treat the information in a layered fashion: the bottom layer is a static network representation, formed by the line and stop\_place entities, the middle layer are the journeys made by vehicles that make stops at specific times, while the top layer are the trips made by the users over these vehicle journeys. In the Figure~\ref{fig:example_network} we show an example for the two bottom layers, where Line 1R is the reversed of Line 1. Our proposed implementation can be found in Section~\ref{sec:ps}.

Finally, it is possible to introduce a \textbf{user} identity, with an anonymized identifier to segregate trips by users. However, we do not consider such information useful for the kind of analysis that this work focuses on. If needed in the future, this additional entity could be trivially integrated in our representation.

@@

\begin{enumerate}[A)]
    \item Aggregation of passengers, asking for the gross number of users that boarded or alighted within an area and a time window. Furthermore, it can be also interesting to obtain the average load of a bus or a train between any two stops from its line. Some of those queries are:
    \begin{itemize}
        \item \texttt{board\_X$_{LT}$} Number of times that stop X is boarded, optionally restricting to a line L and a time range T.
        \item \texttt{alight\_X$_{LT}$} Number of times that stop X is alighted, optionally restricting to a line L and a time range T.
        \item \texttt{board\_L$_T$}  Number of times that line L is boarded, optionally restricting to a time range T.
        \item \texttt{alight\_L$_T$} Number of times that line L is alighted, optionally restricting to a time range T.
        \item \texttt{board\_T} Number of times a user boards within a time range T.
        \item \texttt{alight\_T} Number of times a user alights within a time range T.
        \item \texttt{load\_X$_{LT}$} The average number of passengers traveling from the stop X to its next stop in the line L within the time range T. Can also be seen as the average load of a transport vehicle.
    \end{itemize}
    \item Movement patterns, or queries that focus on the sequence of trip stages. With his kind of queries we can obtain the number of times a stop was used to switch lines or the number of trips that started on a stop with another specific stop as the final destination. In this work we consider the following queries of this kind:
    \begin{itemize}
        \item \texttt{start\_X$_{LT}$} User trips starting at a stop X, optionally restricting to a line L and a time range T.
        \item \texttt{end\_X$_{LT}$} User trips ending at a stop X, optionally restricting to a line L and a time range T.
        \item \texttt{switch\_X$_{LT}$} Number of trips in which the stop X was used to switch lines, optionally restricting to a line L and a time range T.
        \item \texttt{from\_X$_{LT}$\_to\_Y$_{LT}$} User trips that originate at stop X to end at stop Y, both being optionally restricted to a line and time range.
        \item \texttt{start\_L$_T$} User trips starting at any stop from the line L, optionally restricting to a time range T.
        \item \texttt{end\_L$_T$} User trips ending at any stop from the line L, optionally restricting to a time range T.
        \item \texttt{start\_T} User trips starting within a time range T.
        \item \texttt{end\_T} User trips ending within a time range T.
    \end{itemize}
\end{enumerate}

\section{Proposed Structures}
\label{sec:ps}
As of today, we are not aware of any indexing structure that would allow to efficiently represent trajectories that could also support all the kinds of queries described in Section~\ref{sec:rq}. For this reason, we propose two data structures, \acummname~(\acumm)~and~\ttctrname~(\ttctr). The former is targeted for queries of type A, solving most aggregation queries in constant time, while the later can be used for queries of type B. Finally, we introduce a more versatile alternative to \ttctr~that we call \ctrname~(\ctr).

\subsection{Common Data Structures}
\label{sec:cs}
Both \acumm~and \ttctr~need a network representation to be operated, along with information about the transport vehicle journeys. Therefore, the following structures represent a network with stops $s_i \in S = \{1,...|S|\}$, lines $l_i \in L = \{1,...|L|\}$ and journeys $j_i \in J^l = \{1,...|J^l|\}$. It is important to state that journeys are \textbf{not} identified by $j_i$, as the same $j_i$ can belong to several $J^l$ from different lines, so we speak about journey \textbf{codes} (jcodes) instead of journey identifiers.

@@

\subsection{\acumm}
\label{sec:acumm}
As it has been stated previously in this paper, nowadays it could be relatively simple to gather massive information about user trips in every public transport; however, it is not necessary to store individually each trip. If there are two (or more) commuters sharing the same bus (train/tram/etc), they are also sharing the route, the stops (the time of each stop) and even the network. Therefore, a structure storing the aggregated data of the user trips would be a handy solution to solve (at least) queries of type A presented in section 3.

@@

Thus, we propose the T-Matrix, an accumulated solution that involves two-dimensional matrices of integers enabling aggregated queries by row, column, or window/range. In the context of a public transportation network, it would be useful to solve flexible line-centered queries; that is, queries that support aggregation by any dimension. Therefore, it could be possible to aggregate either by time-interval (e.g. number of users got on at any stop of the line on 2019/01/02); by stop (e.g. number of commuters that got on a bus at stop S); or by stop and time-interval.

@@

These weaknesses motivated us to develop a second version, \ctr, which reduces the complexity of the queries that restrict the final line, and delegates line checks on a new WM, allowing for more space-time trade-off. As in \ttctr, the input trips need to be sorted by the same criteria, but in \ctr~we use three complementary structures to represent each component of the sequence, as shown in Figure~\ref{fig:example_xctr}:
\begin{enumerate}[(i)]
    \item An adapted Compressed Suffix Array (\texttt{CSA}) over the stop identifiers of all trips, concatenated into a string with additional terminator symbols $\$$ appended at the end of each trip. As in the \texttt{CSA} from \ttctr, we make these $\$$ symbols maintain the order of the trips and cyclical in $\Psi$. Because this time we do not encode line and stop identifiers together and \texttt{CSA} only encodes stops, there is no need for a complex vocabulary anymore.
    \item \texttt{WML}: Aligned to the entries of (i) there is a Wavelet Matrix (WM) for the line identifiers of each stop. Aligned to the $\$$ section we duplicate the starting lines of each trip. As a trivial optimization, we build a separate WM for every stop, allowing us to save space due to the fact that a single stop does not usually belong to many lines, thus the average height of these WM is no larger (and usually smaller) than the height of a single WM.
    \item \texttt{WMJ}: A WM of jcodes aligned to the leaves of (ii). Note that this makes this structure dependant on (ii), which is coherent with the fact that journey codes themselves are relative to the line identifier. In case (ii) implements the optimization described, the entries of the WM must also be rearranged to match the delimited stops.
\end{enumerate}

@@

>>>>

\usepackage{enumerate}
\usepackage{marginnote}
\usepackage[linesnumbered,commentsnumbered,boxed]{algorithm2e}
\usepackage{amsmath}
\usepackage{threeparttable}
\usepackage{multirow}
\usepackage[margin=0pt,skip=1pt]{subcaption}
\usepackage{xcolor}
\usepackage{fixltx2e}

\newcommand{\acumm}{T-Matrices} % T is for Total!
\newcommand{\acummname}{\mathsf{Trip-Matrices}}
\newcommand{\ctr}{XCTR}
\newcommand{\ctrname}{\mathsf{eXtended~Compact~Trip~Representation}}
\newcommand{\ttctr}{TTCTR}
\newcommand{\ctrq}{$\alpha$}
\newcommand{\ttctrq}{$\beta$}
\newcommand{\ttctrname}{\mathsf{Topology\&Trip-aware~Compact~Trip~Representation}}

\definecolor{red}{rgb}{1.0, 0.0, 0.0}

\newcommand{\ojo}[1]{{\color{red} OJO: \tiny #1}}

\begin{document}
%
\title{New Trippy Structure \thanks{Funded in part by European Union`s Horizon 2020 research and innovation programme
under the Marie Sklodowska-Curie grant agreement No 690941 (project
BIRDS). The Spanish group is also partially funded by Xunta de Galicia/FEDERUE
[CSI: ED431G/01; GRC: ED431C 2017/58; GEMA:IN852A
2018/14 and Innovapeme: IN848D-2017-2350417]; by MINECO-AEI/FEDER-UE
[Datos 4.0: TIN2016-78011-C4-1-R; and ETOMERDFD3:
TIN2015-69951-R]; and by FPI Program [BES-C-2017-0085]. M. A. Rodr´ıguez is
partially funded by Fondecyt-Conicyt grant number 1170497 and by the Millennium
Institute for Foundational Research on Data. }}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Nieves R. Brisaboa\inst{1} \and
Antonio Fari\~na\inst{1}\and
Daniil Galaktionov\inst{1,2} \and Tirso V. Rodeiro\inst{1} \and M. Andrea Rodr\'iguez\inst{3,4}}
%
\authorrunning{N. Brisaboa et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Universidade da Coru\~na, Laboratorio de Bases de Datos, A Coru\~na, Spain  \email{\{brisaboa, fari, d.galaktionov, tirso.varela.rodeiro\}@udc.es}  \and 
Enxenio S.L., A Coru\~na, Spain
\and
Computer Science Department, Universidad de Concepci\'on, Concepci\'on, Chile \and Millennium Institute for Foundational Research on Data, Santiago, Chile
\email{andrea@udec.cl }\\
 }
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
The abstract should briefly summarize the contents of the paper in
15--250 words.

@@

A practical representation for this information in a data structure that supports efficient indexed search would have numerous possible applications, specially for transportation companies interested in studying the ridership \cite{tu2018spatial} and, therefore, improve the efficiency of the network, such as in \cite{weng2018mining}. Another application that is enabled by such analysis is the targeted advertising \cite{zhang2017targeted}, as being able to predict travel patterns for different user profiles could be extremely useful for an advertiser.

@@

\subsubsection{Outline}
This work is structured as follows. We start with a detailed review of related works in Section~\ref{sec:rw}, where we discuss the state of the art for trajectory indexing, mining and then we present the underlying structures that our representation is based on. It is followed by Section~\ref{sec:model}, where we explain our design for a data model that can support relevant queries over transportation networks. In Section~\ref{sec:ps} we detail the composition of our representations and analyze their time efficiency for the targeted queries, followed by a practical analysis in Section~\ref{sec:exp}, where we evaluate the performance on a real transportation network. Finally, the conclusions can be found in Section~\ref{sec:conc}.

\section{Related works}
\label{sec:rw}
\subsection{Trajectory indexing}
Literature on spatial trajectory indexing can be categorized by the nature of the trajectories: they can be either constrained to a network or in free space. While there are well-known queries for indexes that work on moving objects in free space \cite{DBLP:conf/vldb/PfoserJT00}, the network constrained trajectory indexes cover more diverse querying needs, as different networks involve different kinds of challenges (as in vehicular road network vs public transportation network), and also because the intended application may vary (analyzing trip-chaining patterns vs number of passengers within a time window). A comprehensive review on indexing methods can be found in \cite[Chapter 4]{DBLP:books/sp/PelekisT14}, to which we shall expand in the rest of this section to mention the most notable indexes and some new developments \ojo{Desde la referencia NO}.

\subsubsection{Free trajectory indexing}
Several adaptations of the {\em R-Tree} \cite{DBLP:conf/sigmod/Guttman84} are widely used for the indexing of moving objects. The most common approach is to integrate the temporal dimension in the R-Tree itself, as found in the {\em STR-Tree} and the {\em TB-Tree} from \cite{DBLP:conf/vldb/PfoserJT00}.
Another common approach is to complement the R-Tree with another similar structure for time representation, as has been done for the {\em MV3R-tree}~\cite{DBLP:conf/vldb/PapadiasT01},
where an Historical {\em R-Tree}~\cite{nascimento1998towards} to partition on the temporal dimension.

@@

Recently a framework based on Apache Spark called {\em UlTraMan}~ \cite{ding2018ultraman} was developed. It supports different kinds of partitioning schemes for large collections of trajectories to answer range, kNN or aggregation queries \ojo{Queries NO definidas aun}, allowing to divide the dataset to maximize the query time efficiency for a given query type. \ojo{Nieves: No claro} Although UlTraMan has been tested over network constrained trajectories, it does not appear to be exploiting network information in any way, hence its inclusion in this category.

In the field of compact data structures, an index called {\em GraCT}~\cite{brisaboa2019gract} has been developed, based on snapshots at regular time intervals representing the real position of the object at that time using {\em k$^2$-Trees}~\cite{brisaboa2009k} and movement logs for individual trajectories to which grammar compression (using {\em RePair}~\cite{larsson2000off}) is applied. Because of this, GraCT is a self-indexed compact representation that supports spatio-temporal range and nearest-neighbor queries, as well as allowing for direct access to the trajectory information of objects moving in open space (planes, ships\dots).

\subsubsection{Network constrained trajectory indexing}
There are also numerous approaches that use R-Tree based indexes for trajectories that are constrained to an underlying network, aiming to decrease the redundancy in the representation by separating the representation into levels. \ojo{Nieves: no claro} Examples of such indexes include the {\em FNR-Tree}~\cite{DBLP:conf/ssd/Frentzos03}, where the network elements are indexed with a 2D R-Tree and 1D R-Trees to index time intervals of moving objects \ojo{NO} along the edges of the network. Some of the limitations of the FNR-Tree are addressed in \cite{DBLP:journals/geoinformatica/AlmeidaG05}, where the authors propose the {\em MON-Tree}, where the moving objects are indexes in two dimensional R-Trees (the dimensions being edge position vs time). More recent alternatives, such as the {\em TMN-Tree}~\cite{chang2010tmn}, integrate {\em B$^+$-Trees}, which have proven to be more space and time efficient for indexing the temporal dimension. Alternatively, in \cite{rivera2018faster} a compact representation of time intervals is proposed using two bitvectors, that can be combined with these R-Tree based indexes to increase the efficiency for temporal filtering. Refer to \cite{john2017performance} for a quick comparison of these R-Tree based indexes.

@@

Another relevant representation is described in \cite{DBLP:conf/gis/KroghPTT14}. Their proposed index, {\em NETTRA}, was designed to efficiently solve a specific kind of query called {\em Strict Path Queries (SPQ)} that ensure that selected trajectories not only touch with the searched path but also follow it from its beginning to its end. Built on a traditional RDBMS with $B^+$-Tree indexes. Another distinctive feature of NETTRA is its treatment of network constrained trajectories as textual information, which allows to apply string matching techniques such as fingerprinting functions to determine what trajectories have similar paths on their traversed edges. This close equivalence between trajectories and strings has been further exploited by Geodabs~\cite{chapuis2018geodabs}, where both the spatial distribution and sequence information are taken into account for finding trajectories by similarity with fingerprinting. These text-based approaches are sometimes tangled with works on the topic of semantic trajectories such as \cite{al2017semantictraj}.

@@

A recent compact data structure named {\em CiNCT} has been proposed in \cite{koide2018cinct}, where trajectories are encoded into a string, that is used to build an FM-index \cite{DBLP:conf/focs/FerraginaM00} with a Huffman-shaped Wavelet Tree \cite{ferragina2009compressed}. Due to space optimization, the string is constructed with relative movement labels instead of absolute edges, with an auxiliary structure that represents a network graph built from the input trajectories themselves. While this structure excels at pattern-matching and path extraction in street networks, it cannot be applied to our problems in public transportation networks. \ojo{¿Por que?}

\subsection{Trajectories retrieval}
Currently there are several known techniques that would allow to collect data regarding users' trips over a public transportation network. Numerous works exist where those trajectories are mined from the transactions of smart cards \cite{bhaskar2015passenger,wang2014aggregated}. This can be complemented with information derived from GPS devices, as shown in \cite{ma2014development}. Alternatively, reliable trajectories may be extracted relying on positioning obtained from cellular networks, as proven by \cite{liu2017exploring}.

@@

This summarized review proves that \ojo{¿La nuestra?}, although we were not able to access real data from a public transportation company for this work, such curated information about users' trips can indeed be obtained and used for our proposed representations.

\subsection{Underlying structures and algorithms}
\subsubsection{Summed Area Tables}
The Summed Area Tables were first introduced in computer graphics \cite{crow1984summed} to speed up the mipmapping process. In this work we use this approach in order to manage aggregated data about user trajectories based on the idea that an amount of passengers share the same ride (with all its implications: times, trajectories...).

\subsubsection{Bitvectors}
\label{sec:bit}
A great amount of works in Compressed Data Structures have proposed solutions based on bitvectors, both compressed and uncompressed. For a bitvector $B[1,n]$ the following operations are expected to be supported:

\begin{itemize}
    \item $rank_1(B,i)$ returns the number of set bits in $B[1..i]$. Alternatively, $rank_0(B,i) = i - rank_1(B,i)$ and also $B[i] = rank_1(B,i-1) - rank_1(B,i)$. \ojo{¿Que aporta?}
    \item $select_1(B,i)$ returns the position in $1..n$ where the $i$th 1 occurs. Therefore, $rank_1(B,select_1(B,i)) = i$.
\end{itemize}

@@

\subsubsection{Compressed Suffix Array (CSA)}
\label{sec:csa}
One common structure to count or locate all the occurrences of a pattern $P[1,m]$ in a string $S[1,n]$ of symbols from an alphabet $\Sigma$ is the Suffix Array (SA) \cite{MM93}. The SA lexicographically sorts all the suffixes $\{S[i..n] : 1 \leq i \leq n\}$ and builds an array $A[1,n]$ with their positions in the \ojo{???}, holding that $S[A[i]..n] \prec S[A[i+1]..n]$. This allows to find out all the occurrences of a substring $P[1,m]$ in $O(m \log n)$ time with binary searches.

A simple enhancement to avoid storing the original string $S$ is to align $A$ with a bitvector $D[1,n]$ where $D[i]=1~\Leftrightarrow~S[A[i-1]] \prec S[A[i]]$. That is, $D$ marks with a 1 all the positions in $A$ where the starting symbol of that suffix $S[A[i]]$ has a larger lexicographical value than the one from the previous positions $S[A[i-1]]$. With $D$, keeping $S$ is not longer needed as $S[A[i]] = rank_1(D,i)$. \ojo{Nieves no entiende}
%, where $rank_1(D,i)$ returns the number of set bits in $D[1..i]$.

@@

\section{Our model for compact trip representation}
\label{sec:model}
In our previous structures introduced in \cite{brisaboa2018new}, we propose a model to represent the demand information for a public transportation network, as seen in the Figure~\ref{fig:er}. \ojo{Este parrafo regu}

@@

This approach allows us to treat the information in a layered fashion: the bottom layer is a static network representation, formed by the line and stop\_place types, the middle layer represents the journeys made by vehicles that make stops at specific times, while the top layer are the trips made by the users over these vehicle journeys. In the Figure~\ref{fig:example_network} we show an example for the two bottom layers, where Line 1R is the reversed of Line 1. Our proposed implementation can be found in Section~\ref{sec:ps}.

Finally, it is possible to introduce a \textbf{user} identity, with an anonymized identifier to split trips by users. However, we do not consider such information useful for the kind of analysis that this work focuses on. If needed in the future, this additional entity could be trivially integrated in our representation.

@@

\begin{enumerate}[A)]
    \item Queries about the network load, asking for the gross number of users that boarded or alighted within a stop and a time/journey. Furthermore, it can be also interesting to obtain the average load of a bus or a train between any two stops from its line. Some of those queries are:
    \begin{itemize}
        \item \texttt{board\_X$_{LT}$} Number of users that stop X is boarded, optionally restricting to a line L and a time range T.
        \item \texttt{alight\_X$_{LT}$} Number of users that stop X is alighted, optionally restricting to a line L and a time range T.
        \item \texttt{board\_L$_T$}  Number of users that line L is boarded, optionally restricting to a time range T.
        \item \texttt{alight\_L$_T$} Number of users that line L is alighted, optionally restricting to a time range T.
        \item \texttt{board\_T} Number of users boarding within a time range T.
        \item \texttt{alight\_T} Number of users alighting within a time range T.
        \item \texttt{load\_X$_{LT}$} The average number of passengers traveling from the stop X to its next stop in the line L within the time range T. Can also be seen as the average load of the vehicle.
    \end{itemize}
    \item Queries about user trips patterns. With his kind of queries we can obtain the number of times a stop was used to switch lines or the number of trips that started on a stop with another specific stop as the final destination. In this work we consider the following queries of this kind:
    \begin{itemize}
        \item \texttt{start\_X$_{LT}$} User trips starting at a stop X, optionally restricting to a line L and a time range T.
        \item \texttt{end\_X$_{LT}$} User trips ending at a stop X, optionally restricting to a line L and a time range T.
        \item \texttt{switch\_X$_{L\textsubscript{1}L\textsubscript{2}T}$} Number of trips in which the stop X was used to switch lines, optionally restricting to a line L\textsubscript{2} and a time range T.
        \item \texttt{from\_X$_{LT}$\_to\_Y$_{LT}$} User trips that originate at stop X to end at stop Y, both being optionally restricted to a line and time range.
        \item \texttt{start\_L$_T$} User trips starting at any stop from the line L, optionally restricting to a time range T.
        \item \texttt{end\_L$_T$} User trips ending at any stop from the line L, optionally restricting to a time range T. \ojo{revisar}
        \item \texttt{start\_T} User trips starting within a time range T.
        \item \texttt{end\_T} User trips ending within a time range T.
    \end{itemize}
\end{enumerate}

\section{Proposed Structures}
\label{sec:ps}
To the best of our knowledge, there is no indexing structure that would allow to efficiently represent trajectories that could also support all the kinds of queries described in Section~\ref{sec:rq}. For this reason, we propose two data structures, \ojo{UNA SOLA SOLUCION} \acummname~(\acumm)~and~\ttctrname~(\ttctr). The former is targeted for queries of type A, solving most aggregation queries in constant time, while the later can be used for queries of type B. Finally, we introduce a more versatile alternative to \ttctr~that we call \ctrname~(\ctr).

\subsection{Common Data Structures}
\label{sec:cs}
\ojo{TRATAR COMO UNA SOLA SOLUCION} Both \acumm~and \ttctr~need a network representation to be operated, along with information about the transport vehicle journeys. Therefore, the following structures represent a network with stops $s_i \in S = \{1,...|S|\}$, lines $l_i \in L = \{1,...|L|\}$ and journeys $j_i \in J^l = \{1,...|J^l|\}$. It is important to state that journeys are \textbf{not} identified by $j_i$, as the same $j_i$ can belong to several $J^l$ from different lines, so we speak about journey \textbf{codes} (jcodes) instead of journey identifiers.

@@

\subsection{\acumm}
\label{sec:acumm}
As it has been stated previously in this paper, nowadays it could be relatively simple to gather massive information about user trips in every public transport; however, it is not necessary to store individually each trip. If there are two (or more) traveler sharing the same bus (train/tram/etc), they are also sharing the route, the stops (the time of each stop) and even the network. Therefore, a structure storing the aggregated data of the user trips would be a handy solution to solve (at least) queries of type A presented in section 3.

@@

Thus, we propose the T-Matrix, an accumulated solution that involves two-dimensional matrices of integers enabling aggregated queries by row, column, or window/range. In the context of a public transportation network, it would be useful to solve flexible line-centered queries; that is, queries that support aggregation by any dimension. Therefore, it could be possible to aggregate either by time-interval (e.g. number of users got on at any stop of the line on 2019/01/02); by stop (e.g. number of passengers that got on a bus at stop S); or by stop and time-interval.

@@

These weaknesses motivated us to develop a second version, \ctr, which reduces the complexity of the queries that restrict the final line, and delegates line checks on a new WM, allowing a better space-time trade-off. As in \ttctr, the input trips need to be sorted by the same criteria, but in \ctr~we use three complementary structures to represent each component of the sequence, as shown in Figure~\ref{fig:example_xctr}:
\begin{enumerate}[(i)]
    \item An adapted Compressed Suffix Array (\texttt{CSA}) over the stop identifiers of all trips, concatenated into a string with additional terminator symbols $\$$ appended at the end of each trip. As in the \texttt{CSA} from \ttctr, we make these $\$$ symbols maintain the order of the trips and cyclical in $\Psi$. Because this time we do not encode line and stop identifiers together and \texttt{CSA} only encodes stops, there is no need for a complex vocabulary anymore.
    \item \texttt{WML}: Aligned to the entries of (i) there is a Wavelet Matrix (WM) for the line identifiers of each stop. Aligned to the $\$$ section we duplicate the starting lines of each trip. As a trivial optimization, we build a separate WM for every stop, allowing us to save space due to the fact that a single stop does not usually belong to many lines, thus the average height of these WM is no larger (and usually smaller) than the height of a single WM.
    \item \texttt{WMJ}: A WM of jcodes aligned to the leaves of (ii). Note that this makes this structure dependant on (ii), which is coherent with the fact that journey codes themselves are relative to the line identifier. In case (ii) implements the optimization described, the entries of the WM must also be rearranged to match the delimited stops.
\end{enumerate}

@@

    \ojo{Consultas sin sentido}
